### 2010-04-01 papers/icfp-2010-desc/paper_desc.tex
Desc@ICFP: bring in AllI and allI
Desc@ICFP: align codes, whenever possible
Desc@ICFP: fix colors
start 3 run away by 10

In this section, we describe an implementation of inductive types, as
we know them in ML-like languages. By working with datatypes we are
familiar with, we rely on our intuition on these objects. However, our
goal remains to discuss datatypes as we find them in dependently-typed
languages.

Our proposal is inspired by Dybjer and Setzer's closed formulation of
induction-recursion~\cite{dybjer:axiom-ir}. For inductive datatypes, the
full power of induction-recursion is not necessary. Therefore, we
adopt a stripped-down version of it.

@@

In dependently-typed languages, $\Sigma$-types can be interpreted in
two different ways. This duality is actually reflected in the notation
we can find in the literature. The notation $\blue{\Sigma}_{\V{x} :
  \V{A}} (\V{B}\: \V{x})$ stresses that $\Sigma$-types are a
generalisation of sums over arbitrary arities.  Where simply-typed
languages have finite sums, dependently-typed languages have sums
indexed by any set.

On the other hand, our choice of notation $\SIGMA{\V{x}}{\V{A}}
(\V{B}\:\V{x})$ emphasises that $\Sigma$-types are a generalisation of
products, where the second component can depend on the first
one. Where simply-typed languages pack data into non-dependent tuples,
we can now build tuples where data influence further types.

@@

In ML-like languages, datatypes are presented as a
\emph{sum-of-products}. A datatype is defined by a finite sum of
constructors, each composed by a product of arguments. To embrace
these datatypes, we have to capture this grammar.

With dependent types, the notion of sum-of-products translates into
\emph{sigmas-of-sigmas}. However, a direct encoding of datatypes
through raw $\Sigma$-types is not a viable option. Indeed, to program
with these datatypes, we need an induction principle. An encoding is
unable to deliver such
operation~\cite{geuvers:induction-not-derivable}.

\subsection{A universe of descriptions}
\label{sec:desc-universe}

@@

While sigmas-of-sigmas form a convenient \emph{semantics} for
datatypes, we have seen that it makes a poor \emph{syntax}. To
decouple these two notions, we construct an ad-hoc
\emph{universe}. This technique originates in Martin-L\"of definition
of type theory~\cite{martin-lof:itt}. It is a fruitful programming
technique in dependently-typed
languages~\cite{benke:universe-generic-prog, oury:power-of-pi}.

The key idea behind universe construction is our ability to make names
by defining new types. These names are called \emph{codes}. By
defining a set of codes, we define the syntax of a language. To give
it a semantics, we \emph{interpret} these codes back into the type
theory. Hence, codes act as labels, while the type theory provides the
computational machinery. Codes being mere labels, we can inspect them,
hence regaining structural information. We can also compute over them:
deriving new codes from others, or even functions on them.

@@

To embed inductive types in our type theory, we define a universe of
\emph{descriptions} (Figure~\ref{fig:desc_universe}). As expected, the
code of this universe mimics the grammar of inductive types. To build
sigmas-of-sigmas, we define a $\DSigma{\!}{}$ code, interpreted as a
$\Sigma$-type. Descriptions are terminated by $\DUnit$, containing no
payload. To make recursive arguments, we introduce the $\DIndx{\!}$
code. It interprets to a hole that will be closed by the fixpoint. We
can also read $\V{D}$ as describing an endofunctor on $\Set$, with
$\descop{\V{D}}{\!}$ being the object part.

\begin{figure}

\[\stk{
\begin{array}{ll}
\stk{
\data \Desc : \Set \where \\
\;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
    \DUnit          & \Desc \\
    \DSigma{\!}{\!} & \PI{\V{S}}{\Set} \PIS{\V{S} \To \Desc} \Desc \\
    \DIndx{\!}      & \Desc \To \Desc
\end{array}
}
\vspace{0.2in}
\\
\stk{
\descop{\_\:}{} : \Desc \To \Set \To \Set \\
\begin{array}{@{}ll@{\:=\:\:}ll}
\descop{\DUnit}{& \V{X}}        &  \Unit                                       \\
\descop{\DSigma{\V{S}}{\V{D}}}{& \V{X}} &  \SIGMAS{\V{s} : \V{S}}{\descop{\V{D}~\V{s}}{\V{X}}}         \\
\descop{\DIndx{\V{D}}}{& \V{X}}     &  \TIMES{\V{X}}{\descop{\V{D}}{\V{X}}}
\end{array}
}
\end{array}
}\]

\caption{Universe of Descriptions}
\label{fig:desc_universe}

\end{figure}

@@

\[
\begin{array}{lcl}
\F{induction} & : & \PITEL{\V{D}}{\Desc}
                    \PI{\V{P}}{\Mu{\V{D}} \To \Set}         \\
              &   & (\PI{\V{xs}}{\descop{\V{D}}{(\Mu{\V{D}})}}              \\
              &   & \ \All{\V{D}}{(\Mu{\V{D}})}{\V{P}}{\V{xs}} \To \V{P} (\Con{\V{xs}})) \To \\
              &   & \PI{\V{x}}{\Mu{\V{D}}} \V{P} \V{x}
\end{array}
\]

Where, intuitively, $\All{\V{D}}{\V{X}}{\V{P}}{\V{x}}$ states that
$\Bhab{\V{P}}{\V{X} \To \Set}$ holds everywhere in $\V{x}$. Its
definition is presented in Figure~\ref{fig:all-predicates}.  We refer
the reader to our technical report for the implementation of
$\F{induction}$. Note that this induction principle is the first
manifestation of a generic operation over descriptions. Any datatype
we define automatically comes with an induction principle.

@@

\[
\begin{array}{ll}
%%
\stk{
\begin{array}{@{}ll}
\All{\!}{\!}{\!}{\!} : & \PITEL{\V{D}}{\Desc}
                         \PITEL{\V{X}}{\Set}
                         \PITEL{\V{P}}{\V{X} \To \Set} \\
                       & \PI{\V{xs}}{\descop{\V{D}}{\V{X}}} 
                         \Set 
\end{array} \\
\All{\DUnit}{\V{X}}{\V{P}}{\Void} = 
    \Unit \\
\All{(\DSigma{\V{S}}{\V{T}})}{\V{X}}{\V{P}}{\pair{\V{a}}{\V{b}}{}} = 
    \All{(\V{T}\: a)}{\V{X}}{\V{P}}{\V{b}} \\
\All{(\DIndx{\V{D}})}{\V{X}}{\V{P}}{\pair{\V{a}}{\V{b}}{}} =
    \TIMES{\V{P}\: \V{a}}{\All{\V{D}}{\V{X}}{\V{P}}{\V{b}}} \\
\All{(\DHindx{\V{H}}{\V{D}})}{\V{X}}{\V{P}}{\pair{\V{f}}{\V{b}}{}} =
    \TIMES{(\PI{\V{i}}{\V{H}} \V{P}\: (\V{f}\: \V{i}))}
          {\All{\V{D}}{\V{X}}{\V{P}}{\V{b}}}
}
&
%%
\stk{
\begin{array}{@{}ll}
\all{\!}{\!}{\!}{\!}{\!} : & \PITEL{\V{D}}{\Desc}
                             \PITEL{\V{X}}{\Set}
                             \PITEL{\V{P}}{\V{X} \To \Set} \\
                           & \PITEL{\V{R}}{\PI{\V{x}}{\V{X}} \V{P}\: \V{x}}
                             \PI{\V{xs}}{\descop{\V{D}}{\V{X}}} 
                             \All{\V{D}}{\V{X}}{\V{P}}{\V{xs}} 
\end{array} \\
\all{\DUnit}{\V{X}}{\V{P}}{\V{R}}{\Void} =
    \Void \\
\all{(\DSigma{\V{S}}{\V{T}})}{\V{X}}{\V{P}}{\V{R}}{\pair{\V{a}}{\V{b}}{}} =
    \all{(\V{T}\: a)}{\V{X}}{\V{P}}{\V{R}}{\V{b}} \\
\all{(\DIndx{D})}{\V{X}}{\V{P}}{\V{R}}{\pair{\V{a}}{\V{b}}{}} =
    \pair{\V{R}\: \V{a}}
         {\all{\V{D}}{\V{X}}{\V{P}}{\V{R}}{\V{b}}}{} \\
\all{(\DHindx{H}{D})}{\V{X}}{\V{P}}{\V{R}}{\pair{\V{f}}{\V{b}}{}} =
    \pair{\LAM{i} \V{R}\: (\V{f}\: \V{i})}
         {\all{\V{D}}{\V{X}}{\V{P}}{\V{R}}{b}}{}
\end{array}
}
\]

>>>>

In this section, we describe an implementation of inductive types, as
we know them in ML-like languages. By working with familiar datatypes,
we hope to focus on the delivery mechanism, warming up gently to the
indexed datatypes we really want.  % Our encoding is inspired by
Dybjer and Setzer's closed formulation of
induction-recursion~\cite{dybjer:axiom-ir}, but without the
`-recursion'.  An impredicative Church-style encoding of datatypes is
not adequate for dependently typed programming, as although such
encodings present data as non-dependent eliminators, they do not
support dependent \emph{induction}~\cite{geuvers:induction-not-derivable}.
Whilst the \(\LAMBINDER\)-calculus captures all that data can \emph{do},
it cannot ultimately delimit all that data can \emph{be}.

@@

In dependently typed languages, $\Sigma$-types can be interpreted as
two different generalisations. This duality is reflected in the
notation we can find in the literature. The notation
$\blue{\Sigma}_{\x : \M{A}} (\M{B}\: \x)$ stresses that
$\Sigma$-types are `dependent sums', generalising of sums over
arbitrary arities, where simply typed languages have finite sums.

On the other hand, our choice of notation $\SIGMA{\x}{\M{A}}
(\M{B}\:\x)$ emphasises that $\Sigma$-types generalise products,
with the type of the second component depending on the value of the
first, where simply typed languages do not express such relative validity.

@@

In ML-like languages, datatypes are presented as a
\emph{sum-of-products}. A datatype is defined by a finite sum of
constructors, each carrying a product of arguments. To embrace
these datatypes, we have to capture this grammar.
With dependent types, the notion of sum-of-products translates into
\emph{sigmas-of-sigmas}.

\subsection{A universe of datatype descriptions}
\label{sec:desc-universe}

@@

While sigmas-of-sigmas can give a \emph{semantics} for the
sum-of-products structure in each node of the tree-like values in a
datatype, we need to account somehow for the recursive structure which
ties these nodes together. Not for the first time, we do this by
constructing a \emph{universe}~\cite{martin-lof:itt}. Universes
are ubiquitous in dependently typed
programming~\cite{benke:universe-generic-prog, oury:power-of-pi},
but here we seek to exploit them as the foundation of our notion
of datatypes.

%The key idea behind universe construction is our ability to make names
%by defining new types. These names are called \emph{codes}. By
%defining a set of codes, we define the syntax of a language. To give
%it a semantics, we \emph{interpret} these codes back into the type
%theory. Hence, codes act as labels, while the type theory provides the
%computational machinery. Codes being mere labels, we can inspect them,
%hence regaining structural information. We can also compute over them:
%deriving new codes from others, or even functions on them.

@@

To add inductive types to our type theory, we build a universe of
datatype \emph{descriptions} by implementing the signature below, with
codes mimicking the grammar of datatype declarations. We
can read a description $\V{D}:\Desc$ as an \emph{endofunctor}
on $\Set$, with $\descop{\V{D}}{\!}$ its action on an object, \(\V{X}\), soon
to be instantiated recursively.
\[\stk{
\begin{array}{l}
\Desc : \Set\\
\DUnit : \Desc\\
\DSigma{(\Bhab{\V{S}}{\Set})}{(\Bhab{\V{D}}{\V{S} \To \Desc})} : \Desc\\
\DIndx{(\Bhab{\V{D}}{\Desc})} : \Desc \\
\end{array}
\smallskip\\
\begin{array}{l@{\V{X}\:\mapsto\:}l}     
\multicolumn{2}{l}{\descop{\_\:}{} : \Desc \To \Set \To \Set} \\
 \descop{\DUnit}{} &  \Unit \\
 \descop{\DSigma{\V{S}}{\V{D}}}{} &
     \SIGMAS{\Bhab{\V{s}}{\V{S}}}{\descop{\V{D}\,\V{s}}{\!\V{X}}}  \\
\descop{\DIndx{\V{D}}}{}  &  \TIMES{\V{X}}{\descop{\V{D}}{\V{X}}}
\end{array}
}\]
Descriptions are sequential structures, terminated by $\DUnit$, indicating
the empty tuple. To build
sigmas-of-sigmas, we define a $\DSigma{\!}{}$ code, interpreted as a
$\Sigma$-type. To request a recursive component, we have $\DIndx{\V{D}}$,
where \(\V{D}\) descibes the rest of the node.

@@

\[\stk{
\begin{array}{lcl}
\F{induction} & : & \PITEL{\V{D}}{\Desc}
                    \PI{\V{P}}{\Mu{\V{D}} \To \Set}         \\
              &   & (\PI{\V{xs}}{\descop{\V{D}}{(\Mu{\V{D}})}}              \\
              &   & \ \All{\V{D}}{(\Mu{\V{D}})}{\V{P}}{\V{xs}} \To \V{P} (\Con{\V{xs}})) \To \\
              &   & \PI{\V{x}}{\Mu{\V{D}}} \V{P} \V{x} 
\end{array} \\
\F{induction}\: \V{D}\: \V{P}\: \V{ms}\: (\Con{\V{xs}}) =  \\
\qquad
    \V{ms}\: \V{xs}\: (\all{D}{\Mu{D}}{P}
                           {(\LAM{\V{x}} \F{induction}\: \V{D}\: \V{P}\: \V{ms}\: \V{x})}
                           {\V{xs}})
}\]

Where, intuitively, $\All{\V{D}}{\V{X}}{\V{P}}{\V{x}}$ states that
$\Bhab{\V{P}}{\V{X} \To \Set}$ holds everywhere in $\V{x}$. Its
definition is presented in Figure~\ref{fig:all-predicates}. Note that
this induction principle is the first manifestation of a generic
operation over descriptions. Any datatype we define automatically
comes with an induction principle.

@@

\[
\begin{array}{ll}
%%
\stk{
\begin{array}{@{}ll}
\All{\!}{\!}{\!}{\!} : & \PITEL{\V{D}}{\Desc}
                         \PITEL{\V{X}}{\Set}
                         \PITEL{\V{P}}{\V{X} \To \Set} \\
                       & \PI{\V{xs}}{\descop{\V{D}}{\V{X}}} 
                         \Set 
\end{array} \\
\begin{array}{@{}l@{}l@{\:=\:\:}l}
\All{\DUnit}{& \V{X}}{\V{P}}{\Void} &
    \Unit \\
\All{(\DSigma{\V{S}}{\V{T}})}{& \V{X}}{\V{P}}{\pair{\V{a}}{\V{b}}{}} &
    \All{(\V{T}\: \V{a})}{\V{X}}{\V{P}}{\V{b}} \\
\All{(\DIndx{\V{D}})}{& \V{X}}{\V{P}}{\pair{\V{a}}{\V{b}}{}} &
    \TIMES{\V{P}\: \V{a}}{\All{\V{D}}{\V{X}}{\V{P}}{\V{b}}} \\
\All{(\DHindx{\V{H}}{\V{D}})}{& \V{X}}{\V{P}}{\pair{\V{f}}{\V{b}}{}} &
    \TIMES{(\PI{\V{i}}{\V{H}} \V{P}\: (\V{f}\: \V{i}))}
          {\All{\V{D}}{\V{X}}{\V{P}}{\V{b}}}
\end{array}
}
&
%%
\stk{
\begin{array}{@{}ll}
\all{\!}{\!}{\!}{\!}{\!} : & \PITEL{\V{D}}{\Desc}
                             \PITEL{\V{X}}{\Set}
                             \PITEL{\V{P}}{\V{X} \To \Set} \\
                           & \PITEL{\V{R}}{\PI{\V{x}}{\V{X}} \V{P}\: \V{x}}
                             \PI{\V{xs}}{\descop{\V{D}}{\V{X}}} 
                             \All{\V{D}}{\V{X}}{\V{P}}{\V{xs}} 
\end{array} \\
\begin{array}{@{}l@{}l@{\:=\:\:}l}
\all{\DUnit}{& \V{X}}{\V{P}}{\V{R}}{\Void} &
    \Void \\
\all{(\DSigma{\V{S}}{\V{T}})}{& \V{X}}{\V{P}}{\V{R}}{\pair{\V{a}}{\V{b}}{}} &
    \all{(\V{T}\: \V{a})}{\V{X}}{\V{P}}{\V{R}}{\V{b}} \\
\all{(\DIndx{\V{D}})}{& \V{X}}{\V{P}}{\V{R}}{\pair{\V{a}}{\V{b}}{}} &
    \pair{\V{R}\: \V{a}}
         {\all{\V{D}}{\V{X}}{\V{P}}{\V{R}}{\V{b}}}{} \\
\all{(\DHindx{\V{H}}{\V{D}})}{& \V{X}}{\V{P}}{\V{R}}{\pair{\V{f}}{\V{b}}{}} &
    \pair{\LAM{\V{i}} \V{R}\: (\V{f}\: \V{i})}
         {\all{\V{D}}{\V{X}}{\V{P}}{\V{R}}{\V{b}}}{}
\end{array}
\end{array}
}
\]

### 2010-04-01 papers/icfp-2010-desc/paper_type_theory.tex
start 3 run away by 10

Dually, type checking judgments
(Fig.~\ref{fig:type-checking}) are \emph{sinks} for types. From an
$\exprIn$ and a type pushed into it, they elaborate a low-level
term, extracting information from the type.
Canonical set-formers are \emph{checked}: we could exploit \(\Set:\Set\)
to give them synthesis rules, but this would prejudice our future
stratification plans. Note that abstraction and pairing are free of
annotation, as promised. Most of the propagation rules are unremarkably
structural: we have omitted some mundane rules which just follow
the pattern, e.g., for \(\UId\). It is for the finite enumerations that
things get interesting.

@@

Finally, we enable you to write functions from finite sets as tuples.
Any bracketed form, \(\void\) or \(\pair{\_}{\_}{}\), is accepted by the
function space---the general product---if it is accepted by the
small product. Propagation fills in the appeal to the $\F{switch}$,
copying the range information.

We note, but have no room to elaborate on the fact that the propagation
is reversible. Indeed, our interactive development tools rely on the
reverse transformation to relay intelligible information to the user.
For any specific usage, the encoding is completely hidden by these
translations. Only when programming generically with enumerations is it
necessary to address the encoding, but in that case, it is a benefit!

@@

The relatively civilised presentation we have achieved here is
possible because our universe of enumerations specifies not only the
\emph{representation} of the low-level values in each set as bounded
numbers, but also the \emph{presentation} of these values as
high-level tags. To encode only the former, we should merely need the
\emph{size} of enumerations, but we extract more work from these types
by making them more informative.  We have also, \emph{en passant},
distinguished enumerations which have the same cardinality but
describe distinct notions:
\(\EnumT{\sqr{\etag{\CN{red}}\,\etag{\CN{blue}}}}\) is not
\(\EnumT{\sqr{\etag{\CN{green}}\,\etag{\CN{orange}}}}\).

>>>>

Dually, type checking judgments
(Fig.~\ref{fig:type-checking}) are \emph{sinks} for types. From an
$\exprIn$ and a type pushed into it, they elaborate a low-level
term, extracting information from the type. Note that we inductively ensure the following `sanity checks':
\[\stkc{
\Gamma\Vdash\propag{e}{\pull{t}{T}} \Rightarrow
  \Gamma\vdash t:T \\
\Gamma\Vdash\push{\propag{e}{t}}{T} \Rightarrow
  \Gamma\vdash t:T
}\]

Canonical set-formers are \emph{checked}: we could exploit
\(\Set:\Set\) to give them synthesis rules, but this would prejudice
our future stratification plans. Note that abstraction and pairing are
free of annotation, as promised. Most of the propagation rules are
unremarkably structural: we have omitted some mundane rules which just
follow the pattern, e.g., for \(\UId\). It is for the finite
enumerations that things get interesting.

@@

Finally, we let you express functions from enumerations as tuples.
Any tuple-form, \(\void\) or \(\pair{\_}{\_}{}\), is accepted by the
function space---the general product---if it is accepted by the
small product. Propagation fills in the appeal to $\F{switch}$,
copying the range information.

We note, but have no room to elaborate on the fact that the propagation
is reversible. Our interactive development tools rely on the
reverse transformation to report intelligibly to the user.
The encoding of any specific enumeration is thus hidden by these
translations. Only in enumeration-generic programs is the
encoding exposed, but in that case, it is a benefit!

@@

Our type propagation mechanism does no fancy unification or constraint
solving, just copying.  The relative civilision it delivers is
possible because our universe of enumerations specifies not only the
\emph{representation} of the low-level values in each set as bounded
numbers, but also the \emph{presentation} of these values as
high-level tags. To encode only the former, we should merely need the
\emph{size} of enumerations, but we extract more work from these types
by making them more informative.  We have also, \emph{en passant},
distinguished enumerations which have the same cardinality but
describe distinct notions:
\(\EnumT{\sqr{\etag{\CN{red}}\,\etag{\CN{blue}}}}\) is not
\(\EnumT{\sqr{\etag{\CN{green}}\,\etag{\CN{orange}}}}\).
